Data Specialization for a Staged Lambda Calculus

### 2015-02-13 paper/splitting.tex

In this section, we provide an algorithm to translate the multistage terms and values of \lang\
into equivalent but simpler forms where the stage \bbone\ and stage \bbtwo\ portions 
of the computation are {\em syntactically separated} from each other.
That is, we desire outputs of the form $\mathcal{C}(e_1;e_2)$ where  
$e_1$ is a monostage term encoding all stage \bbone\ computation, 
and $e_2$ is a monostage term encoding all stage \bbtwo\ computation.
To represent outputs in this way, we introduce three new constructs:
\begin{itemize}
\item {\em Single-output pipelines}, \texttt{pipeS}, for the output of splitting \lang\ terms at world \bbtwo.
\item {\em Multi-output pipelines}, \texttt{pipeM}, for the output of splitting \lang\ terms at world \bbonem. 
\item {\em Masked values}, \texttt{mv}, which will serve a purpose analogous to \lang\ partial values.
\end{itemize}

We also define analogs to evaluation and residualization for these new terms, 
called $\sepredonesym$ and $\sepredtwosym$, 
which respectively send \texttt{pipeM} terms to \texttt{mv} terms
and \texttt{pipeS} terms to residuals.
An overview of these features is provided in \ref{fig:outputTermSummary}.
That figure also provides the actual rules for $\sepredonesym$ and $\sepredtwosym$.
Note how there's one rule for each, and they're rather simple.
Indeed, the simplicity of these reduction rules is one reason why we 
would want to translate into this separated form, rather than use the semantics of \ref{sec:semantics}.

With these output terms defined, we can state the goal of our splitting algorithm:
to translate terms in \lang\ to a syntactically-separated form, 
such that the semantics is preserved.

Exactly what is meant by ``such that the semantics is preserved'' depends on whether we're translating
terms at \bbonem\ or \bbtwo.  
In the latter case, where an $\coltwo e A$ is translated to a $\pipeS p l r$ with the $\splittwosym$ relation,
this constraint is straightforward: $e$ and $\pipeS p l r$ can both be reduced to residuals
via $\redtwosym$ and $\sepredtwosym$, so we only need to require that those
residuals be equivalent for the translation to be correct.
In the former case, where an $\colmix e A$ is translated to a $\pipeM c l r$ with the $\splitonesym$ relation, 
things are more complicated: 
the output of $\redonesym$ is a partial value and the output of $\sepredonesym$ is a masked value,
so we first need a way to relate those forms if we want to talk about correctness of the term translation.
That is the purpose of the {\em masking} relation, written $\vsplito$.

This section proceeds by defining masking ($\vsplito$), 
then using masking to motivate the definition of \bbonem-translation ($\splitonesym$),
and finally coming back to define \bbtwo-translation ($\splittwosym$).
A summary of these operations and the relationship between them is provided in \ref{fig:splittingSummary}.

>>>>

In this section, we provide a {\em splitting algorithm} to translate the multistage terms of \lang\,
into equivalent but simpler forms where the stage \bbone\ and stage \bbtwo\ portions 
of the computation are {\em syntactically separated} from each other.

More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
we can translate $e$ into some syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$), which
can also be reduced to the residual $q$, up to an equivalence.
We enforce the ``syntactically separated'' condition structurally by saying that 
$\mathcal{P}$ must have the form $\pipeS p l r$ where  
$p$ (called the {\em precomputation}) is a monostage term encoding all the stage \bbone\ computation that was in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term encoding all the stage \bbtwo\ computation that was in $e$.  
We can then residualize $\mathcal{P}$ with the $\sepredtwosym$ judgement, which is defined by the single rule
which evaluates the precomputation and plugs it in for $l$:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}),
we also provide a way to split those as well.
This is given by the new relation $e \splitonesym \mathcal{P}$,
where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
$c$ (called the {\em combined term}) is a monostage term encoding all the stage \bbone\ computation that was in $e$, 
and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the stage \bbtwo\ computation that was in $e$.

Similar to above, we also define a $\sepredtwosym$ relation which can be used to evaluate \texttt{pipeM} terms,
with one complication arising from the fact that \lang\ terms at world \bbonem\ produces multistage output.
For example, consider the term
\begin{lstlisting}
(next {1+2}, pure{3+4})
\end{lstlisting}
which reduces (via $\redonesym$) to the residual table and partial value
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, pure{7})
\end{lstlisting}
Since this output has interleaved stages, so theres no way a syntactically seperated term could directly reduce to it.

We resolve this mismatch by introducing one more translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
which converts residual tables and partial values like the those above into a syntactically seperated version $\mathcal{V}$.
As before, we enforce the ``syntactically seperated'' property structurally,
saying that $\mathcal{V}$ must have the form $\mval i q$ (called a {\em masked value}) where 
$i$ is a monostage value respresenting the stage \bbone\ part of $v$,
and $q$ is a monostage term representing all of the stage \bbtwo\ computation in $\xi$ and $v$.

With this machinery in place, we can define $\sepredtwosym$ as just creating masked values with the single rule:
\[
\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
Here we see the primary difference between terms at world \bbtwo\ and world \bbone.
Since $\coltwo e A$ reduces to an entirely stage \bbtwo\ result, 
its stage \bbone\ subcomputations only exist to internally pass their value to stage \bbtwo;
but since $\colmix e A$ reduces to a multistage result, 
its stage \bbone\ subcomputations exist {\em both} to pass their value to stage \bbtwo\ {\em and} to be present in the result.
Correspondingly, in the $\sepredtwosym$ rule, $p$ reduces only to a boundary value which is passed to the resumer;
but in the $\sepredonesym$ rule, $c$ reduces to a tuple containing both the immediate result and the boundary value,
with only the latter being passed to the resumer.

A summary of these operations and the relationship between them is provided on the left side of \ref{fig:splittingSummary}.

This section proceeds by defining masking ($\vsplito$), 
then using masking to motivate the definition of \bbonem-translation ($\splitonesym$),
and finally coming back to define \bbtwo-translation ($\splittwosym$).
